---
title: "Exploratory Data Analysis for EU ETS"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  arning = FALSE, 
  message = FALSE,
  warning = FALSE)
source('function_vs.R')
```

```{r}
# Load library
library("tidyverse")
library("kableExtra")   # R markdown tables
library("paletteer")  # Nice figure color 
library("scales")
library("eurostat")  # EU maps
library("sf")
library("giscoR")
```

```{r}
results <- read_csv('../01_tidy_data/ets_merged.csv')
```


This is an R notebook that documents the EDA for the EU ETS.

# Data Overview

The table below summarises the number of unique observations for different variables. There are 17,281 \* 17 = 302,957 unique installation-year.

```{r}
results %>%
  select(account_holder_id, acc_id, installation_id, com_year) %>%
  rename(
    "Account Holder ID" = account_holder_id,
    "Account ID" = acc_id,
    "Installation ID" = installation_id,
    "Compliance Year" = com_year
  ) %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), names_to = "Variables", values_to = "Unique Values") %>%
  kable(align = "lc") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Regulated Entities


The EU ETS covers both the installations and aircraft operators, with the latter being regulated starting from 2012. The two figures below show the number of regulated entities in each year for installations and aircraft operators, respectively.

\vspace{15mm}

```{r, fig.width=6,fig.height=2.75}
ggplot(
  data = results %>%
    filter(!ins_isAircraftOperator & regulated) %>%
    group_by(com_year) %>%
    summarise(count = n(), .groups = "drop"), 
  aes(x = com_year, y = count)
) +
  geom_bar(
    stat = "identity", 
    fill = "darkslategrey", 
    color = "black"
  ) +
  labs(
    title = "Number of Regulated Installations",
    x = "Year",
    y = "Count"
  ) +
  scale_x_continuous(
    breaks = seq(2005, 2022, by = 2)
  ) +  
  scale_y_continuous(
    breaks = pretty_breaks(n = 8)
  ) +   
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12)
  )
```



```{r, fig.width=6,fig.height=2.75}
ggplot(
  data = results %>%
    filter(ins_isAircraftOperator & regulated) %>%
    group_by(com_year) %>%
    summarise(count = n(), .groups = "drop"),
  aes(x = com_year, y = count)
) +
  geom_bar(
    stat = "identity", 
    fill = "lightblue", 
    color = "black"
  ) +
  labs(
    title = "Number of Regulated Aircraft Operators",
    x = "Year",
    y = "Count"
  ) +
  scale_x_continuous(
    breaks = seq(2012, 2021, by = 1)
  ) +  
  scale_y_continuous(
    breaks = pretty_breaks(n = 8)
  ) +   
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12)
  )
```

\vspace{15mm}

The plot below shows the number of ever regulated installations in each country.

```{r}
# Obtain EU map data
eu_shp <- get_eurostat_geospatial(
  resolution = 10, 
  nuts_level = 0, 
  year = 2016) %>%
  # Rename the ID for Greece from EL to GR for matching
  mutate(id = case_when(
    id == "EL" ~ "GR",
    id == "UK" ~"GB",
    TRUE ~ id))
```

```{r}
eda_EUmap(
  data = eu_shp %>%
    inner_join(results %>%
               distinct(installation_id, ins_country_id) %>%
               group_by(ins_country_id) %>%
               summarise(Count = n()),
               by = join_by(id == ins_country_id)) %>%
    arrange(geo) %>% 
    st_as_sf(),
  fill_var = "Count",
  breaks = seq(0, 3000, by = 500),
  name = "No. of Installations",
  title = "Number of Ever Regulated Installations by Country",
  subtitle = "From 2005 to 2021",
  caption = "Data: EUTL"
)
```


\newpage

The figure below shows the regulated entities by EU ETS regulated activities, regrouped by activity category. The most common activity is the activity involving combustion of fuels.

\vspace{10mm}

```{r, fig.width=6,fig.height=3}
# Add activity category to results
results <- results %>%
  mutate(category = case_when(
      ins_activity_id %in% c(1, 20) ~ "Combustion", 
      ins_activity_id %in% c(2, 21) ~ "Refineries", 
      ins_activity_id %in% c(3, 22) ~ "Coke ovens",
      ins_activity_id %in% c(4, 5, 23:28) ~ "Metal",
      ins_activity_id %in% c(6, 29, 30) ~ "Cement and Lime",
      ins_activity_id %in% c(7, 8, 31:34) ~ "Glass and Ceramics",
      ins_activity_id %in% c(9, 35, 36) ~ "Pulp and Paper",
      ins_activity_id == 10 ~ "Aircrafts",
      ins_activity_id %in% c(37:42, 44) ~ "Chemicals",
      ins_activity_id == 43 ~ "Hydrogen Production",
      ins_activity_id %in% c(45:47) ~ "Carbon Capture and Storage",
      ins_activity_id == 99 ~ "Other"))


ggplot(
  data = results %>% 
    distinct(installation_id, ins_activity_id, category) %>% 
    group_by(category) %>% 
    summarise(Count = n()) %>%
    mutate(category = reorder(category, Count)),
  aes(x = category, y = Count)
) +
  geom_bar(
    stat = "identity", 
    fill = "deepskyblue4", 
    color = "black"
  ) +
  labs(
    title = "Number of Ever-Regulated Entities by Activity Category",
    x = "Activity Category",
    y = "Count"
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 9)
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12)
  ) +
  coord_flip()
```

# Emissions
This section will explore the emission and compliance of the EU ETS. The figure below shows the aggregated verified emissions and aggregated free allocated allowances across all entities by year.

\vspace{10mm}

```{r, fig.width=6,fig.height=3.5}
bar_data <- results %>%
  group_by(com_year) %>% 
  summarise(total = sum(com_verified_sum, na.rm = TRUE)/1e6, .groups = "drop")

line_data <- results %>%
  group_by(com_year) %>% 
  summarise(total = sum(com_allocatedTotal_sum, na.rm = TRUE)/1e6, .groups = "drop")

ggplot() +
  # Bar chart for verified emissions
  geom_bar(
    data = bar_data,
    aes(x = com_year, y = total),
    stat = "identity", 
    fill = "coral4", 
    color = NA,
    alpha = 0.8
  ) +
  # Line chart for allocated allowances
  geom_line(
    data = line_data,
    aes(x = com_year, y = total),
    color = "darkolivegreen",
    size = 1.2
  ) +
  # Points on the line
  geom_point(
    data = line_data,
    aes(x = com_year, y = total),
    color = "black",
    size = 2.5
  ) +
  # Labels and theme
  labs(
    title = "Aggregated Verified Emissions and Free Allocated\nAllowances (MtCO2eq) by Year",
    x = "Year",
    y = "MtCO2eq",
    caption = "Bar: Verified Emissions | Line: Free Allocated Allowances"
  ) +
  scale_x_continuous(
    breaks = seq(2005, 2022, by = 2)
  ) +  
  scale_y_continuous(
    breaks = pretty_breaks(n = 8)
  ) +   
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12)
  )

rm(bar_data, line_data)
```

\vspace{10mm}

The figures below show the average verified emissions by different 

```{r, fig.width=6,fig.height=3.5}
eda_activity_emission(
  data = results, 
  selected_categories = c(
    "Refineries", 
    "Coke ovens",
    "Metal", 
    "Cement and Lime"
  ),
  palette_name = "beyonce::X124"
)
```

\vspace{15mm}

```{r, fig.width=6,fig.height=3.5}
eda_activity_emission(
  data = results, 
  selected_categories = c(
    "Combustion",
    "Chemicals",
    "Aircrafts",
    "Hydrogen Production"
  ),
  palette_name = "lisa::OskarSchlemmer"
)
```
```{r, fig.width=6,fig.height=3.5}
eda_activity_emission(
  data = results, 
  selected_categories = c(
    "Glass and Ceramics",
    "Pulp and Paper",
    "Carbon Capture and Storage",
    "Other"),
  palette_name = "lisa::FridaKahlo"
)


```







